---
title: "Data Analysis: Group 16"
author: "2889642, 2888928, 2960701, 2896766"
number-sections: true
theme: flatly
format:
  pdf:
    titlepage: true
    toc: false
    geometry: margin=1in
classoption: titlepage
execute:
  echo: false
  message: false
  warning: false
editor: visual
header-includes:
  - \usepackage{float}
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(tidyverse)
library(sjPlot)
library(performance)
library(kableExtra)
```

# Introduction

In 1998, an article was published by *Andrew Wakefield* and 12 others that falsely claimed that the measles, mumps, and rubella (MMR) vaccine might be linked to autism. This caused widespread public concern, leading to many worried parents worldwide to stop vaccinating their children, which put many children at a higher risk of catching the disease. The article was later retracted after scientists proved there was no link between the MMR vaccine and autism.

This report analyses data on measles susceptibility among children in 101 intermediate zones (IZs) in Edinburgh between 1998 and 2012. The dataset records the year the data was collected, the total number of pre-school children $(N)$, and the number of those children who were susceptible to measles $(Y)$.

The main questions of interest are whether measles susceptibility in Edinburgh changed following the articles retraction, and whether any such change occurred specifically in 2004, the year the retraction took place.

@sec-eda consists of an exploratory analysis of the measles data, focusing on the distribution of susceptibility rates over time to identify trends and potential shifts surrounding the 2004 retraction. @sec-fda contains the results from fitting a model to the data, as well as the assessment of the model assumptions. Concluding remarks are given in @sec-con.

# Exploratory Data Analysis {#sec-eda}

```{r}
#| echo: false
#| message: false

measles <- read.csv("DAProject16.csv") %>% 
  mutate(prop = Y/N)
```

To appropriately compare measles susceptibility across different intermediate zones, we must account for varying population sizes. Therefore, we calculate the susceptibility proportion $(Y/N)$ for each IZ.

To investigate whether Edinburgh exhibited a change in measles susceptibility following the retraction of the Wakefield article, @tbl-summary displays the mean, median, and standard deviation of the susceptibility rate across the observed years.

```{r}
#| echo: false
#| label: tbl-summary
#| tbl-cap: Mean, median and standard deviation (sd) of the measles susceptibility proportion by Year.


summary <- measles %>%
  summarise('Prop_Mean' = mean(prop, na.rm = TRUE),
            'Prop_Median' = median(prop, na.rm = TRUE),
            'Prop_sd' = sd(prop, na.rm = TRUE),
            .by = Year)

summary %>%
  kable(col.names = c("Year",
                      "Mean Proportion",
                      "Median Proportion",
                      "Proportion SD"),
        digits = 3)
```

From the article's release in 1998 to its retraction in 2004, the mean proportion of children susceptible to measles dropped from $0.095$ to $0.080$. The mean proportion continued to drop after the article was retracted falling to $0.050$ in 2012.

@fig-box illustrates the overall distribution and spread of these susceptibility rates.

```{r}
#| echo: false
#| fig-cap: Distribution of measles susceptibility rates across intermediate zones from 1998 to 2012
#| label: fig-box
#| fig-align: center
#| fig-width: 4
#| fig-height: 2.5
#| fig-pos: 'H'

ggplot(measles, aes(x = factor(Year), y = prop)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(x = "Year of Data Collection",
       y = "Susceptibility Proportion") +
  theme_minimal()
```

The median of susceptibility proportion decreases steadily from the article's release in 1998 to its retraction. After the article was retracted the median continues to decrease. The variability of the susceptibility proportion is much lower in 2010 and 2012 compared to the other years of data collection.

To determine if the change in measles susceptibility occurred exactly in 2004 alongside the articles retraction, @fig-trend visualises the mean susceptibility rate over the entire time period. The dashed vertical line highlights 2004 so we can observe whether the rate drops immediately during this year, or if the decline is delayed.

```{r}
#| echo: false
#| fig-cap: Trend of the mean measles susceptibility rate over time. The dashed line marks the 2004 retraction. 
#| label: fig-trend
#| fig-align: center
#| fig-width: 4
#| fig-height: 2.5
#| fig-pos: 'H'
ggplot(summary, aes(x = Year, y = Prop_Mean)) +
  geom_line(colour = "slateblue4", linewidth = 1, aes(group = 1)) +
  geom_point(colour = "slateblue4", size = 3) +
  geom_vline(xintercept = 2004, linetype = "dashed", colour = "springgreen4", linewidth = 1) +
  scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2008, 2010, 2012)) +
  labs(x = "Year of Data Collection",
       y = "Mean Susceptibility Proportion") +
  theme_minimal()
```

It is clearly seen in the above figure that the mean susceptibility proportion drops significantly in the years following the articles retraction. The biggest change in mean susceptibility proportion comes from 2008 to 2010 with the steepest line segment in @fig-trend.

To further show if there was a change in the proportion of children susceptible to measles @fig-box2 shows a box plot comparing the data by article status.

```{r}
#| echo: false
#| fig-cap: Distribution of measles susceptibility rates across intermediate zones from 1998 to 2012
#| label: fig-box2
#| fig-align: center
#| fig-width: 4
#| fig-height: 2.5
#| fig-pos: 'H'

ggplot(measles, aes(x = ifelse(Year < 2004, "1998-2003", "2004-2012"), y = prop)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(x = "Year of Data Collection",
       y = "Susceptibility Proportion") +
  theme_minimal()
```

@fig-box2 shows that the median number of children susceptible to measles falls after the article was retracted. After the article was retracted the variability of the susceptibility proportion was lower. There are a few outliers after the articles retraction which may require further investigation.

# Formal Data Analysis {#sec-fda}

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(broom)
library(kableExtra)

data <- read.csv("DAProject16.csv")

data <- data %>%
  mutate(
    proportion = Y/N,
    Year_c = Year - mean(Year),
    post2004 = ifelse(Year >= 2004,1,0)
  )

```

To formally investigate measles susceptibility, we model the number of susceptible children (Y) out of the total number of children (N) using a binomial generalised linear model with a logit link.

For each observation, $$
Y_{i} \sim \text{Binomial}(N_{i},\, p_{i})
$$ where $p_{i}$ represents the probability that a child in a given intermediate zone and year is susceptible to measles.

The quantity of interest is therefore the proportion of susceptible children: $$
p_{i} = \frac{Y_i}{N_i},
$$

which is modelled as a function of time using logistic regression.

## Model 1: Temporal trend

The first model examines whether measles susceptibility changes over time. The logistic regression model is specified as

$$
\text{logit}(p_{i}) = \beta_0 + \beta_1 \cdot \text{Year}_{i},
$$

where:

-   $p_{i}$ is the probability of susceptibility

-   $\beta_0$ is the intercept

-   $\beta_1$ represents the temporal trend.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-mdl1
#| tbl-cap: Logistic regression results for temporal trend model
model1 <- glm(cbind(Y, N-Y) ~ Year_c,
  family = binomial,
  data = data)

model1_table <- tidy(model1)

kable(model1_table,
  digits=4)
```

@tbl-mdl1 shows that the coefficient for the year is negative $(\hat{\beta}_1 = -0.049)$ and highly significant $(p < 0.001)$. This indicates that the probability of children being susceptible to measles decreases over time. This suggests that measles immunity in the population improved during the study period.

On the odds scale, $\exp(-0.049) \approx 0.952$ which suggests that the odds of susceptibility decrease by approximately **4.8% per year** on average.

This provides strong evidence that measles susceptibility in Edinburgh has changed over time.

## Model 2: Change after 2004

To investigate whether susceptibility changed after the retraction of the Wakefield article in 2004, we introduce an indicator variable: $$
Post2004 = 
\begin{cases} 
1 & Year \geq 2004 \\ 
0 & otherwise 
\end{cases}
$$ The logistic model becomes: $$
\text{logit}(p_{i}) = \beta_0 + \beta_1 \cdot \text{Year}_{i} + \beta_2 \cdot \text{Post2004}_{i}
$$ where:

-   $\beta_1$ represents the overall temporal trend

-   $\beta_2$ measures any additional change after 2004.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-mdl2
#| tbl-cap: Logistic regression results including post-2004 indicator


model2 <- glm(cbind(Y, N-Y) ~ Year_c + post2004,
  family = binomial,
  data = data)

model2_table <- tidy(model2)

kable(model2_table,
  digits=4)
```

In @tbl-mdl2, the model shows that the overall year effect remained significantly negative$(\hat{\beta}_1 = -0.0645, p < 0.001)$, while the post-2004 indicator was positive $(\hat{\beta}_2 = 0.1743)$ and statistically significant $(p \approx 0.003)$, indicating that including this variable significantly improves the model.

On the odds scale, $\exp(0.1743) \approx 1.19$ which suggests that, after accounting for the overall linear time trend, the odds of measles susceptibility for observations from 2004 onwards are approximately 19% higher than would be expected from the temporal trend alone.

Importantly, this does not imply that susceptibility increases after 2004. The coefficient for Year remains negative, indicating an overall declining trend in susceptibility across the study period. This provides statistical evidence that measles susceptibility did not follow a single linear temporal trend across the study period, but instead experienced an additional structural shift after 2004.

## Model comparsion

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-comparsion
#| tbl-cap: Likelihood ratio test comparing nested models

anova_results <- anova(model1, model2, test="Chisq")

kable(anova_results,
  digits=4)
```

To assess whether including the post-2004 indicator improves the model, in @tbl-comparsion, a likelihood ratio test was performed to compare the two nested models：

$$
\begin{aligned}
H_0 : \beta_2 &= 0, \\
H_1 : \beta_2 &\neq 0
\end{aligned}
$$

The test result was statistically significant $(p = 0.0032)$, indicating that the model including the post-2004 indicator provides a significantly better fit to the data.

This provides statistical evidence that measles susceptibility differed significantly after 2004.

## Visualization of Trends

To further contextualize these statistical findings, we examine the aggregate annual proportions from the logistic model.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-visual
#| fig-width: 4
#| fig-height: 3
#| fig-pos: 'H'
#| fig-cap: "Trend in Measles Susceptibility"

ggplot(data, aes(x = Year, y = proportion)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = FALSE,
              colour = "steelblue") +
  labs(title = "Trend in Measles Susceptibility",
       x = "Year",
       y = "Proportion susceptible") +
  theme_minimal()
```

@fig-visual shows that the fitted logistic regression curve confirms a clear declining trend in measles susceptibility over time.

# Conclusions {#sec-con}

The empirical analysis of measles susceptibility in Edinburgh (1998–2012) provides a rigorous basis for addressing the research objectives.

The results from the temporal trend model confirm a systematic reduction in susceptibility, with the odds declining by approximately 4.8% per annum. This trajectory indicates a sustained improvement in population immunity over the study period. However, the integration of the post-2004 indicator reveals a significant structural shift following the Wakefield article retraction. Although the long-term trend remained downward, the positive coefficient suggests that susceptibility levels were notably higher after 2004 than predicted by a purely linear trend. The likelihood ratio test confirms that this model offers a superior fit, underscoring the 2004 retraction as a pivotal factor in shaping the city's epidemiological landscape.

In terms of future work, further research could benefit from examining regional differences across Edinburgh’s intermediate zones to better understand local variations in measles susceptibility. Additionally, analyzing more detailed data on vaccination uptake and public health interventions would help identify the socio-demographic factors driving changes in immunity levels, particularly following the 2004 retraction.
