---
title: "Data Analysis: Group 16"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
editor_options: 
  chunk_output_type: console
theme: flatly
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(tidyverse)
library(sjPlot)
library(performance)
library(kableExtra)
```

# 1. introduction

The main questions of interest are:

Did Edinburgh exhibit a change in measles susceptibility following the retraction of the Wakefield article? Did the change, if any, in measles susceptibility occur in 2004 alongside the articles’ retraction?

# Exploratory DA

```{r}
#| echo: false
#| message: false

measles <- read.csv("DAProject16.csv") |>
  mutate(prop = Y/N)
```

To appropriately compare measles susceptibility across different intermediate zones, we must account for varying population sizes. Therefore, we calculate the susceptibility proportion $(Y/N)$ for each IZ.

To investigate whether Edinburgh exhibited a change in measles susceptibility following the retraction of the Wakefield article, @tbl-summary displays the mean, median, and standard deviation of the susceptibility rate across the observed years.

(Interpretation of table)

```{r}
#| echo: false
#| label: tbl-summary
#| tbl-cap: Mean, median and standard deviation (sd) of the measles susceptibility proportion by Year.

summary <- measles %>%
  summarise('Prop_Mean' = mean(prop, na.rm = TRUE),
            'Prop_Median' = median(prop, na.rm = TRUE),
            'Prop_sd' = sd(prop, na.rm = TRUE),
            .by = Year)

summary %>%
  kable(col.names = c("Year",
                      "Mean Proportion",
                      "Median Proportion",
                      "Proportion SD"),
        digits = 3)
```

@fig-box illustrates the overall distribution and spread of these susceptibility rates.

(description of boxplot)

```{r}
#| echo: false
#| fig-cap: Distribution of measles susceptibility rates across intermediate zones from 1998 to 2012
#| label: fig-box
#| fig-align: center
#| fig-width: 6
#| fig-height: 4

ggplot(measles, aes(x = factor(Year), y = prop)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(x = "Year of Data Collection",
       y = "Susceptibility Proportion") +
  theme_minimal()
```

To determine if the change in measles susceptibility occurred exactly in 2004 alongside the articles retraction, @fig-trend visualises the mean susceptibility rate over the entire time period. The dashed vertical line highlights 2004 so we can observe whether the rate drops immediately during this year, or if the decline is delayed.

(description of line graph)

```{r}
#| echo: false
#| fig-cap: Trend of the mean measles susceptibility rate over time. The dashed line marks the 2004 retraction. 
#| label: fig-trend
#| fig-align: center
#| fig-width: 6
#| fig-height: 4

ggplot(summary, aes(x = Year, y = Prop_Mean)) +
  geom_line(colour = "slateblue4", linewidth = 1, aes(group = 1)) +
  geom_point(colour = "slateblue4", size = 3) +
  geom_vline(xintercept = 2004, linetype = "dashed", colour = "springgreen4", linewidth = 1) +
  scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2008, 2010, 2012)) +
  labs(x = "Year of Data Collection",
       y = "Mean Susceptibility Proportion") +
  theme_minimal()
```

