---
title: "Project16"
author: "Group16"
format: html
editor: visual
execute:
  echo: false
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

# Formal Data Analysis {#sec-fda}
```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(broom)
library(kableExtra)

data <- read.csv("DAProject16.csv")

data <- data %>%
  mutate(
    proportion = Y/N,
    Year_c = Year - mean(Year),
    post2004 = ifelse(Year >= 2004,1,0)
  )

```

## Model 1: Temporal trend
$$
Y_{it} \sim \text{Binomial}(N_{i},\, p_{i})
$$
$$
\text{logit}(p_{it}) = \beta_0 + \beta_1 \cdot \text{Year}
$$
```{r}
model1 <- glm(
  cbind(Y, N-Y) ~ Year_c,
  family = binomial,
  data = data
)

summary(model1)
```

The coefficient for Year is negative and highly significant (p < 0.001).
This indicates that the probability of children being susceptible to measles decreases over time. This suggests that measles immunity in the population improved during the study period.

## Model 2: Change after 2004
To examine whether susceptibility changed after the retraction of the Wakefield article in 2004, we introduce an indicator variable:
$$
Post2004 = 
\begin{cases} 
1 & Year \geq 2004 \\ 
0 & otherwise 
\end{cases}
$$

The model becomes:
$$
\text{logit}(p_{it}) = \beta_0 + \beta_1 \cdot \text{Year} + \beta_2 \cdot \text{Post2004}
$$
```{r}
model2 <- glm(
  cbind(Y, N-Y) ~ Year_c + post2004,
  family = binomial,
  data = data
)

summary(model2)
```

## Model comparsion
```{r}
anova(model1, model2, test="Chisq")
```

The coefficient for post2004 is positive and statistically significant (p ≈ 0.003).
This suggests that there was a structural change in measles susceptibility following the retraction of the Wakefield article.
Even after accounting for the overall time trend, susceptibility levels differ significantly after 2004.

```{r}
ggplot(data, aes(x = Year, y = proportion)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "glm",
              method.args = list(family="binomial"),
              se = FALSE,
              colour = "blue") +
  labs(
    title = "Trend in Measles Susceptibility",
    x = "Year",
    y = "Proportion susceptible"
  ) +
  theme_minimal()
```

## Formal conclusions
The statistical analysis indicates that measles susceptibility in Edinburgh changed significantly over the study period.

The logistic regression model reveals a strong negative temporal trend, indicating that susceptibility declined over time.

Furthermore, the post-2004 indicator variable is statistically significant, suggesting that a shift in susceptibility occurred following the retraction of the Wakefield article.

These results provide evidence that vaccination behaviour and population immunity may have changed around the time of the article's retraction.


